---
title: 'Assignment #6'
author: "Ethan Jantz"
date: "11/6/2020"
output:
  word_document:
    toc: no
  pdf_document:
    toc: no
  html_document:
    toc: yes
    toc_float: yes
always_allow_html: yes
---

This document is made using [R Markdown](https://rmarkdown.rstudio.com/), a great tool for sharing research in a way that is readable and reproducible. 

## Assignment 6

***

Using chi squared tests, answer the following two questions: 

    1. Do men and women differ in their highest level of education (V161270)? 
        Recategorize education into 4 categories: Less than a High School Degree, High school or High school Equivalent, Some College/Associates Degree, and 4 Year Degree and Higher.
        
    2. Does the proportion of people who support the 2010 health care law (V161113) vary depending on their religion (V161265x)?
        For the sake of this statistical test, combine the three Protestant groups into one category and combine Jewish and "Other Religion" into one category.
        Your categories should then be Protestants, Catholics, Undifferentiated Christian, Other Religion, and Not Religious. 

Explain in full sentences what you are doing and why you are doing it. Interpret the results. Do not only paste R output into the document but instead include it with your description of what the output means. 

***

```{r setup, echo = TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE) # R Markdown options
library(tidyverse); library(gridExtra) # For putting plots side-by-side, trying out a different package this time
library(rmarkdown) # For accessing the functions used to display tables
library(haven) # For reading ANES2016.dta file
library(descr) 

anes <- haven::read_dta("ANES2016.dta") # Reading in the data
```

# Part A

## Select and Recode Variables

First, I'll be selecting and recoding the variables. The select function selects and renames the variables, the filter function removes observations that have inapplicable responses, the mutate function recodes variables. In this case every variable selected is turned into a factor. 
 
```{r select}
selected <- anes %>%
  select(
    "Gender" = V161342,
    "Education" = V161270,
    "Health_Law_Support" = V161113,
    "Religion" = V161265x
  ) %>%
  filter(
    Gender %in% c(1,2),
    Education %in% c(1:90), # 5 respondents responded "other" and then specified "High School Graduate" and that is coded as 90
    Health_Law_Support %in% c(1:3),
    Religion %in% c(1:8)
  ) %>%
  mutate(
    Gender = factor(Gender, labels = c("Male", "Female")),
    Education = case_when(
      Education %in% c(1:8) ~ 1, # no high school
      Education %in% c(9,90) ~ 2, # high school or equivalent
      Education %in% c(10:12) ~ 3, # some college or associates degree
      Education %in% c(13:16) ~ 4, # college degree or higher
    ) %>% 
      factor(labels = c("Less than a High School Degree", "High School Degree or GED", "Some College/Associates Degree", "4 Year Degree and Higher")),
    
    Health_Law_Support = factor(Health_Law_Support, labels = c("Favor", "Oppose", "Neither favor nor oppose")),
    
    Religion = factor(Religion) %>%
      fct_collapse(
        Protestant = c("1", "2", "3"),
        Catholic = "4",
        "Undifferentiated Christian" = "5", 
        Other = c("6", "7"),
        "Not Religious" = "8"
      )
  )
```

## Comparing Education Attainment by Gender

Here I'm running both the CrossTable (part of the *descr* library) and chisq.test (part of base R). I'm passing **chisq = TRUE** to CrossTable() to compare the results to the base R chisq.test function. The results of this test will tell me if there is an association between educational attainment and gender within the sample population.

```{r }
Education_Gender_CTable <- CrossTable(selected$Education, selected$Gender, prop.c = TRUE, prop.r = TRUE,
                               chisq = TRUE, total.r = TRUE, total.c = TRUE, expected = TRUE)
Education_Gender_chisq.test <- chisq.test(selected$Education, selected$Gender)

Education_Gender_CTable
Education_Gender_chisq.test
```

We can see that the results of the tests between the functions are the same. The benefit to running CrossTable is its built-in description of the data. This is powerful, since I can see that there's likely a weak association between gender and educational attainment and not a strong one, but I still need to run another test to say so confidently. I'm using Cramer's V test because both variables are categorical/nominal and differ in the number of categories. 

```{r }
CramerV(selected$Education, selected$Gender)
```

> **Answer**
The p-value for both chi-squared tests was the same (.0225) and significant (< .05), so we can say that there is an association. The result of Cramer's V test shows that this association is weak (.048).

Given the total distribution of gender in the sample (male = .47, female = .52) we can see that female-identifying respondents are only slightly more likely to have higher educational attainment than male-identifying respondents. The largest difference between the two being in the Some college/Associates Degree category.

# Part B

## Comparing Support for Health Care Law and Religious Affiliation

Here is the text detailing the meaning of the health care law question: 

    Do you favor, oppose, or neither favor nor oppose the health care reform law passed in 2010? This law requires all Americans to buy health insurance and requires health insurance companies to accept everyone.
    
```{r }
HealthLawSupport_Religion <- CrossTable(anesselected$Health_Law_Support, selected$Religion)

HealthLawSupport_Religion
```