---
title: 'Assignment #1'
author: "Ethan Jantz"
date: "9/10/2020"
output:
  word_document:
    toc: no
  html_document:
    toc: yes
    toc_float: yes
always_allow_html: true
---

This document is made using [R Markdown](https://rmarkdown.rstudio.com/), a great tool for sharing research in a way that is readable and easily reproducible. 

***

### Assignment 1

Pick 3 variables from the data file that you want to transform into new variables. At least one of the variables has to be a categorical variable. Describe how each is currently coded and provide frequencies for each variable.

Describe how you transform each into a new variable and provide the R Studio syntax used to do the transformation. Provide frequencies for each new variable.

Using a categorical variable with at least 3 response options, provide a frequency for the variable. Then, exclude respondents who gave one of the possible response options and produce frequencies for the remaining groups. Provide the R Studio syntax and output for this.

For this and all other assignments, please provide text describing what you did and present the information in a way that is clear and easy to follow (so that someone could replicate it if necessary). Please do not just use bullet points or R Studio output. 

***

### Setup

```{r setup, echo = TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE) # R Markdown options
library(tidyverse)
library(rmarkdown) # For accessing the functions used to display tables
library(haven) # For reading ANES2016.dta file
library(here) # For easier file referencing. See documentation at https://here.r-lib.org/

anes <- haven::read_dta("ANES2016.dta") # Reading in the data
```

### Exploring the Data

There are a *lot* of variables to pick from. Using the dim() function, we can see the number of rows and columns: `r dim(anes) # The dim function returns the dimensions of an object`

```{r explore}
head(anes) %>% # The head function returns the first 6 rows of an object 
  paged_table() # The paged_table function takes head(anes) and displays it in the document
```

From here, I'm going to reference the ANES Code Book and pick some variables that I'm interested in.

| Variable Name (dta) | Variable Name (code book) | Type of Variable |
|:--------------------|--------------------------:|-----------------:|
|V168023| Estimated family income| Categorical|
|V162139| Importance of reducing deficit| Ordinal|
|V162002| Post: How many programs about 2016 campaign did R watch on TV| Ordinal|
<!-- |V167501| Number of campaign or political signs visible in the structure's window(s)| Ratio| -->
<!-- |V167558| Relative to the other buildings on the road, is the physical condition of the building better, worse, or the same? | Categorical| -->
<!-- |V162134| How much opportunity in America to get ahead| Ordinal| -->

Let's create a new dataframe object that consists of only the selected variables. Here is the result:

```{r subset}
selected <- anes %>% # Here I use the select function to select and rename the variables being studied
  select(
    "FamIncome" = V168023,
    "Deficit" = V162139,
    "TVPrograms" = V162002
    )

selected %>%
  paged_table() # Finally, display the new dataframe
```

### Variable Description

I've selected several variables, and now I'm going to transform and explore them. I'll start with **V168023**. The procedure here will be to explain how each variable is coded and display the frequencies of the variable.

##### **V168023**: Estimated Family Income
This is the categorical variable I will be focusing on. It breaks the respondent down into 7 categories of family income. Here is a list of the categories and their associated variable code:

1.  Under $15,000
2. $15,000 to $29,999
3. $30,000 to $49,999
4. $50,000 to $74,999
5. $75,000 to $99,999
6. $100,000 to $249,999
7. $250,000 or more

Other variable codes include **-9** Refused, **-8** Don't know, and **-1** Inap (inapplicable).

Below is the frequency table for this variable:

```{r echo = FALSE}
selected$FamIncome %>%  
  table()
```

##### **V162139**: Post-Election - Importance of Reducing Deficit
This response, collected after the election, asks the respondent their feelings on the importance of reducing the federal budget deficit. It breaks down into 5 options:

1. Extremely important
2. Very important
3. Moderately important
4. A little important
5. Not at all important

Other variable codes include **-9** Refused, **-8** Don't know, **-7** No post data, **-6** No post-election interview.

Below is the frequency table for this variable:

```{r echo = FALSE}
selected$Deficit %>%  
  table()
```

##### **V162002**: Post-Election - How Many Programs About 2016 Campaign Did R Watch on TV
This response, collected after the election, asks the respondent how many programs about the 2016 campaign they watched on television. It breaks down into 4 responses:

1. None
2. Just one or two
3. Several
4. A good many

Other variable codes include **-7** No post data and **-6** No post-election interview.

Below is the frequency table for this variable:

```{r echo = FALSE}
selected$TVPrograms %>%  
  table()
```

<!-- ##### **V167501**: Number of Campaign Signs Visible in the Structure's Window -->

<!-- ##### **V167558**: Building Condition Relative to Others on Road -->

<!-- ##### **V162134**: Perspective on Opportunity in America -->

### Variable Transformation

In the following chunk I will be removing observations where the variables can't be used for analysis. From there I will factor the FamIncome variable, which will provide labels for the categories. Aftwards I will output the resulting frequency tables for each variable.

```{r transform, echo = TRUE}
selected <- selected %>%
  filter(
    FamIncome >= 0, # All values below 0 are refused, don't know, or inapplicable
    Deficit > 0, # Negative values are inapplicable, don't know, refused, etc
    TVPrograms > 0 # Negative values are innaplicable, don't know, refused, etc
    )

selected$FamIncome <- factor(
    selected$FamIncome,
    labels = c(
      "Under $15,000",
      "$15,000 to $29,999",
      "$30,000 to $49,999",
      "$50,000 to $74,999",
      "$75,000 to $99,999",
      "$100,000 to $249,999",
      "$250,000 or more"
    )
  )
```

##### **Family Income**
```{r echo = FALSE}
selected$FamIncome %>%  
  table()
```

##### **Deficit Sentiment**
```{r echo = FALSE}
selected$Deficit %>%  
  table()
```

##### **Campaign Programs Seen on TV**
```{r echo = FALSE}
selected$TVPrograms %>%  
  table()
```

One thing to note at this point is the downsizing we just performed on this data set. The original dataset had **`r dim(anes)[1]`** observations, while this transformed dataset has **`r round(dim(selected)[1]/dim(anes)[1]*100, 0)`%** of that at **`r dim(selected)[1]`** observations. We could have replaced the unusable values with NA (or null), but if we want to perform more thorough analysis on this dataset we need complete observations across variables.  

### Categorical Exclusion

Finally, let's see what happens to our data when we exclude the responses of respondents who were in the lowest income bracket:

```{r echo = TRUE}
selected <- selected %>%
  filter(
    as.integer(FamIncome) > 1 
  )
```

##### **Family Income**
```{r echo = FALSE}
selected$FamIncome %>%  
  table()
```

##### **Deficit Sentiment**
```{r echo = FALSE}
selected$Deficit %>%  
  table()
```

##### **Campaign Programs Seen on TV**
```{r echo = FALSE}
selected$TVPrograms %>%  
  table()
```

### Conclusion

After all of this data wrangling, here is the final product. Our new ANES dataset, exported into a tidy comma separated values format:

```{r fin}
selected %>%
  paged_table()
```

And since we would likely be using this dataset in the real world, let's export it too.

```{r export}
write.csv(selected, "ANES_selected.csv") # Export the final dataset to the project folder as .csv
```