---
title: 'Assignment #3'
author: "Ethan Jantz"
date: "10/6/2020"
output:
  html_document:
    toc: yes
    toc_float: yes
  word_document:
    toc: no
always_allow_html: true
---

This document is made using [R Markdown](https://rmarkdown.rstudio.com/), a great tool for sharing research in a way that is readable and easily reproducible. 

***

Part A:

    Select two categorical variables.
        Provide relevant descriptive statistics (frequencies, modal response) for these variables and the syntax used to do so.
    Select two continuous (ordinal, interval, or ratio) variables.
        Provide relevant descriptive statistics (mean, median, measures of dispersion).

Part B:

    Choose a dichotomous categorical variable (or simplify categorical variables with more than two categories).
        First, test whether the proportion of respondents giving a particular response is significantly different from 50%.
        Second, test whether the proportion of respondents giving the response is significantly different from 25%.
    Next, choose a continuous (ordinal, interval, or ratio) variable.
        First test whether the mean of the variable is significantly different from zero (0).
        Second, test whether the mean of the variable is significantly different from one (1).
    Provide all R code and text to accompany and describe what you did and the results.


***

```{r setup, echo = TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE) # R Markdown options
library(psych); library(ltm) # psych for describe(), ltm for cronbach.alpha()
library(tidyverse)
library(rmarkdown) # For accessing the functions used to display tables
library(haven) # For reading ANES2016.dta file
library(here) # For easier file referencing. See documentation at https://here.r-lib.org/

anes <- haven::read_dta("ANES2016.dta") # Reading in the data
```

### Part A

```{r select}
selected <- 
  anes %>%
  select(
    "State" = V161015b,
    "Pres_Vote" = V161027,
    "Dem" = V161086,
    "Rep" = V161086
    )
```

In a previous version of this I filtered the data 

#### Categorical Variables

| Variable Code (dta) | Variable Name (code book) | Categories |
|:--------------------|--------------------------:| ----------:|
|**V161015b**| Registration state (all registrations)|  All 50 states + DC and PR|
|**V161027**|  For whom did R vote for President| Hillary Clinton, Donald Trump, Gary Johnson, Jill Stein, Other |

##### Descriptive Statistics

###### Registration State

```{r }
selected %>%  
  filter(
    State > 0
  ) %>%
  select(State) %>%
  table()
```

###### Presidential Candidate

```{r }
selected %>%  
  filter(
    Pres_Vote > 0
  ) %>%
  select(Pres_Vote) %>%
  table()
```

#### Continuous Variables

| Variable Code (dta) | Variable Name (code book) | Range |
|:--------------------|--------------------------:| -----:|
|**V161086**|Feeling Thermometer: Democratic Presidential cand|0 - 100|
|**V161087**|Feeling Thermometer: Republican Presidential cand|0 - 100|

##### Descriptive statistics

###### Feeling Thermometer: Democratic Presidential cand

```{r }
selected %>%  
  filter(
    Dem > 0
  ) %>%
  select(Dem) %>%
  table()
```

###### Feeling Thermometer: Republican Presidential cand

```{r }
selected %>%  
  filter(
    Rep > 0
  ) %>%
  select(Rep) %>%
  table()
```